---
title: "epidemic-modelling-guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{epidemic-modelling-guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(epimodels)
```
## Simulating a Basic SIR Epidemic with Pedigree Structures
This example demonstrates how to initialise parameters for a basic Susceptible-Infectious-Removed (SIR) epidemic, generate a multi-generation hierarchical pedigree, set up randomized exposure groups, and model the transmission dynamics.

For a basic SIR model, the basic reproduction number is calculated as:
$$ R_0=\beta × \text{RP} $$
where $ \\beta $ is the transmission coefficient and $ \\text{RP} $ is the mean infectious (removal) period.

```{r}
# 1. Define model parameters
params <- make_parameters(
  model_type = "SIR",
  dataset = "sir_demo",
  name = "sir_baseline",
  setup = "single",
  sim_new_data = "r",
  group_layout = "random"
)

# 2. Build the population structures and pedigree
pedigree <- make_pedigree(params)
popn <- set_groups(pedigree, params)

# 3. Simulate genetic traits and downstream epidemic transmission
popn <- set_traits(popn, params)
epic_popn <- simulate_epidemic(popn, params)

# 4. Extract parameters and plot survival trajectories
print(get_R0(epic_popn))
plot_model(epic_popn, params)

```

### 2. Setting Up an SEIDR Model with Multi-Trait Genomic Covariances
In this scenario, we set up a compartmental model containing latent/exposed ($ E $) and undetectable ($ I $) states alongside a detectable ($ D $) status. We initialise individual genetic effects (IEs) for susceptibility (`s`), infectivity (`i`), and tolerance (`t`), defining custom additive genetic variances and correlations.

The structural genetic relationships are formalised via the genomic covariance matrix `$ \boldsymbol{\Sigma}_{G} $`:

```R
# 1. Customising genetic variance-covariance matrices
custom_vars <- list(sus = 1.2, inf = 0.8, tol = 0.5, default = 0.2)
custom_cors <- list(si = -0.3, st = 0.15, it = 0.25, default = 0.0)

params <- make_parameters(
  model_type = "SEIDR",
  dataset = "genetic_seidr",
  name = "multi_trait_scen",
  IEs = "sit",               # Susceptibility, Infectivity, Tolerance
  vars = custom_vars,
  cors = custom_cors,
  setup = "large"
)

# 2. Construct popn with relatedness structures
pedigree <- make_pedigree(params)
popn <- set_groups(pedigree, params)

# 3. Sample genetic traits using Cholesky decomposition of the A-matrix
popn <- set_traits(popn, params)
```

---

### 3. Incorporating and Recentring Body Weight as a Phenotypic Fixed Effect
This example shows how to configure fixed environmental effects (such as individual body weight) and apply them to scale individual disease traits log-linearly before running an epidemic.

```R
# 1. Define fixed effects applying to Susceptibility (s), Infectivity (i), and Tolerance (t)
fixed_effects <- list(weight = "sit")

params <- make_parameters(
  model_type = "SEIR",
  dataset = "covariate_modelling",
  name = "weight_fe_scen",
  FEs = fixed_effects,
  use_weight = "log",
  setup = "fb_12_rpw"
)

# 2. Initialise pedigree and assign weights (re-sampled or bespoke)
popn <- make_pedigree(params)
popn <- set_groups(popn, params)
popn <- set_weights(popn, params) # Applies normal/log-normal weight distributions

# 3. Apply the recentred fixed effects to construct final individual phenotypes
popn <- set_traits(popn, params)
popn <- set_phenotypes(popn, params)
```

---

### 4. Simulating Multiple Trials with Nested Trial Structures
When simulating nested architectures (e.g., matching the biological attributes of the Fishboost dataset), you can segment animals of same-family lineages among multiple separated exposure challenges (e.g., Trial 1 and Trial 2), allowing for different baseline transmission levels.

```R
# 1. Initialise nested multi-trial layouts
params <- make_parameters(
  model_type = "SEIR",
  setup = "fb_12_rpw", # Splitting across Trial 1 & Trial 2
  group_layout = "fishboost",
  sim_new_data = "r"
)

# 2. Retrieve structured data allocations
pedigree <- make_pedigree(params)
popn <- set_groups(pedigree, params)

# 3. Observe the trial divisions assigned to individual groups
table(popn$trial, popn$group)
```

---

### 5. Correcting Historical Inoculation Status (Demoting Non-Infective Seeders)
When evaluating real-world context data (such as challenge schedules), individuals classified as "seeders" ($ donor = 1 $) might fail to show signs of positive infection. This custom handling step cleans the data by demoting unresponsive seeders to contact status ($ donor = 0 $).

```R
# 1. Load biological challenge data
raw_popn <- readRDS("fb_data/fb_12.rds")

# 2. Configure filtering rules
params <- make_parameters(
  model_type = "SEIDR",
  setup = "fb_12",
  sim_new_data = "no" # Use existing database
)
params$fix_donors <- c("time", "no_Tsign_survivors")
params$t_demote <- c(20, 80) # Threshold demotions per trial

# 3. Strip inappropriate donors
corrected_popn <- fix_fb_data(raw_popn, params)
```

---

### 6. Exporting High-Performance R Configuration Packages to BICI
The package natively interfaces with BICI, a C++ command-line application used for heavy computational MCMC and sequence inference. This script demonstrates how to generate the corresponding structured `.bici` package directories files programmatically.

```R
# 1. Make parameters and build challenge population
params <- make_parameters(
  model_type = "SIDR",
  dataset = "bici_export",
  name = "hpc_scen_1"
)
popn <- make_pedigree(params) |> set_groups(params) |> set_traits(params)

# 2. Generate configuration directories
# Creates parameters, data matrices, sparse pedigrees and maps
bici_config_list <- generate_bici_script(popn, params)
```

---

### 7. Running BICI Bayesian Inference via C++ Execution Calls
This example sets up MCMC sampling parameters, compiles script configurations, and initiates BICI execution directly from R.

```R
params <- make_parameters(
  model_type = "SEIR",
  dataset = "mcmc_run",
  name = "inference_test",
  algorithm = "pas" # Parallel Tempering / Parallel MCMC
)
params$nsample <- 1e5
params$nchains <- 8

# Setup configs and drop ancient configurations
cleanup_bici_files(params)
bici_script <- generate_bici_script(popn, params)

# Model execution call to underlying compiled bici software
cmd <- with(params, str_glue(
    "mpirun -n {nchains} --oversubscribe ../BICI/bici-Linux {config} {bici_cmd}"
))
system(cmd)
```

---

### 8. Diagnostic Evaluation and Rebuilding BICI Posteriors
Following BICI model calls, parameters are verified using diagnostic criteria like Effective Sample Size (ESS) and Gelman-Rubin convergence criteria ($ \hat{R} $). 

```R
# Rebuild diagnostics, calculate HDI bounds and save updated results
posterior_estimates <- rebuild_bici_posteriors(
  dataset = "mcmc_run",
  name = "inference_test"
)

# Output summary metric lines to screen
msg_pars(posterior_estimates)
```

---

### 9. Flattening Posterior Trace Files for Disease Trajectory Reconstruction
Individual disease progress files (`state_<n>.txt`) represent specific points in time for the disease compartment states. Extracting these trajectories turns file parameters into long-format tidy tables.

```R
# Flatten state files to compute point estimates of latent compartment times
etc_results <- flatten_bici_states(
  dataset = "mcmc_run",
  name = "inference_test",
  bici_cmd = "inf"
)

# Extract unobserved transitions for specific individuals
individual_transitions <- etc_results$popn[id == 42]
print(individual_transitions)
```

---

### 10. Evaluating Posterior Heritability and Genetic Parameters
This example outlines how to read posterior output trace files, compute realized heritability distributions ($ h^{2} = \sigma^{2}_{g} / (\sigma^{2}_{g} + \sigma^{2}_{e}) $), and plot estimated densities against simulated true parameters.

```R
# 1. Plot genetic parameter profiles and posterior densities 
cov_plots <- plot_covariances(
  dataset = "genetic_seidr",
  scen = 1,
  itn = 1,
  ci = "hpdi" # Highest Posterior Density Interval
)

# 2. Access specific genetic components of the model plot
# Displays posterior densities for additive genetic values (G), 
# environmental residuals (E), and phenotypic distributions (P)
cowplot::plot_grid(cov_plots$G, cov_plots$H, ncol = 2)
```